---
title: "Abalone Age Prediction - Complete Analysis"
author: "Harvey"
date: "Dec 11, 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 12, fig.height = 8)
```

# Introduction

This analysis predicts abalone age using linear regression with engineered features. The pipeline includes:

- Data cleaning and imputation
- Feature engineering (ratios, volumes, interactions)
- 38 single models + 5 ensembles (43 total)
- 10-fold cross-validation
- Model comparison and diagnostics

```{r libraries}
library(MASS)
library(tidyverse)
library(caret)
library(glmnet)
library(gridExtra)
library(corrplot)
library(splines)
```

# Part 1: Load Data

```{r load-data}
abalone <- read_csv("~/Library/CloudStorage/Box-Box/Data Science/1. MSDS - Doing Data Science (Bivin Sadler)/Unit 14 and 15 (Group Project)/train2.csv", show_col_types = FALSE)

glimpse(abalone)
```

# Part 2: Data Cleaning

```{r data-cleaning}
# Fix weight inconsistencies (components > whole)
abalone <- abalone %>%
  mutate(
    component_sum = `Shucked Weight` + `Viscera Weight` + `Shell Weight`,
    Weight = if_else(component_sum > Weight, component_sum, Weight)
  ) %>%
  dplyr::select(-component_sum)

# Impute zeros with sex-specific medians
medians_by_sex <- abalone %>%
  filter(Length > 0, Diameter > 0, Height > 0) %>%
  group_by(Sex) %>%
  summarise(
    median_Length = median(Length),
    median_Diameter = median(Diameter),
    median_Height = median(Height),
    .groups = "drop"
  )

abalone <- abalone %>%
  left_join(medians_by_sex, by = "Sex") %>%
  mutate(
    Length = if_else(Length == 0 | is.na(Length), median_Length, Length),
    Diameter = if_else(Diameter == 0 | is.na(Diameter), median_Diameter, Diameter),
    Height = if_else(Height == 0 | is.na(Height), median_Height, Height)
  ) %>%
  dplyr::select(-median_Length, -median_Diameter, -median_Height)

# Biological flags
abalone <- abalone %>%
  mutate(
    bio_flag = case_when(
      Sex == "I" & Age > 6 ~ "Infant_too_old",
      Sex %in% c("M", "F") & Age < 5 ~ "Adult_too_young",
      TRUE ~ "OK"
    )
  )
```

# Part 3: Feature Engineering

```{r feature-engineering}
abalone <- abalone %>%
  mutate(
    # Ratios
    shucked_ratio = `Shucked Weight` / Weight,
    viscera_ratio = `Viscera Weight` / Weight,
    shell_ratio = `Shell Weight` / Weight,
    meat_ratio = (`Shucked Weight` + `Viscera Weight`) / Weight,
    height_diameter_ratio = Height / Diameter,
    diameter_length_ratio = Diameter / Length,
    height_length_ratio = Height / Length,
    
    # Volume
    volume_ellipsoid = (4/3) * pi * (Length/2) * (Diameter/2) * (Height/2),
    volume_cylinder = pi * (Diameter/2)^2 * Height,
    
    # Density
    density = Weight / volume_ellipsoid,
    
    # Surface area
    surface_area = 4 * pi * ((Length * Diameter)^1.6075 / 3)^(1/1.6075),
    
    # Geometric means
    size_mean = (Length * Diameter * Height)^(1/3),
    size_norm = sqrt(Length^2 + Diameter^2 + Height^2),
    weight_mean = (`Shucked Weight` * `Viscera Weight` * `Shell Weight`)^(1/3),
    weight_norm = sqrt(`Shucked Weight`^2 + `Viscera Weight`^2 + `Shell Weight`^2),
    
    # Shell thickness proxy
    shell_thickness_proxy = `Shell Weight` / surface_area,
    
    # Binary flags
    is_infant = if_else(Sex == "I", 1L, 0L),
    is_male = if_else(Sex == "M", 1L, 0L),
    is_female = if_else(Sex == "F", 1L, 0L),
    is_slow_growth = if_else(bio_flag == "Infant_too_old", 1L, 0L),
    is_early_mature = if_else(bio_flag == "Adult_too_young", 1L, 0L),
    
    # Polynomial terms
    size_mean_squared = size_mean^2,
    size_mean_cubed = size_mean^3,
    size_mean_quartic = size_mean^4,
    weight_squared = Weight^2,
    shell_ratio_squared = shell_ratio^2,
    
    # Original interactions
    size_shell_interaction = size_mean * shell_ratio,
    size_shucked_interaction = size_mean * shucked_ratio,
    weight_infant_interaction = Weight * is_infant,
    shell_infant_interaction = shell_ratio * is_infant,
    size_weight_interaction = size_mean * Weight,
    shell_shucked_interaction = `Shell Weight` * `Shucked Weight`,
    density_size_interaction = density * size_mean,
    
    # Enhanced interactions
    size_viscera_interaction = size_mean * viscera_ratio,
    weight_shell_interaction = Weight * shell_ratio,
    weight_shucked_interaction = Weight * shucked_ratio,
    shell_shucked_ratio_interaction = shell_ratio * shucked_ratio,
    infant_size_squared = is_infant * size_mean_squared,
    slow_growth_size = is_slow_growth * size_mean,
    density_infant_interaction = density * is_infant,
    thickness_infant_interaction = shell_thickness_proxy * is_infant,
    
    # Log transforms
    log_size_mean = log(size_mean),
    log_weight = log(Weight),
    log_shell_weight = log(`Shell Weight`),
    log_shucked_weight = log(`Shucked Weight`),
    log_viscera_weight = log(`Viscera Weight`),
    log_density = log(density),
    log_shell_thickness = log(shell_thickness_proxy),
    
    # Sqrt transforms
    sqrt_size_mean = sqrt(size_mean),
    sqrt_weight = sqrt(Weight),
    sqrt_shell_weight = sqrt(`Shell Weight`)
  )

# Handle infinite/NA
abalone <- abalone %>%
  mutate(across(where(is.numeric), ~if_else(is.infinite(.), NA_real_, .))) %>%
  mutate(across(where(is.numeric), ~if_else(is.na(.), median(., na.rm = TRUE), .)))

abalone <- abalone %>%
  mutate(Sex = factor(Sex))
```

# Part 4: Box-Cox Transformation

```{r boxcox}
bc_model <- lm(Age ~ size_mean + Weight + `Shell Weight`, data = abalone)
bc <- boxcox(bc_model, plotit = TRUE)
optimal_lambda <- bc$x[which.max(bc$y)]
cat("Optimal Box-Cox lambda:", round(optimal_lambda, 3), "\n")

if (abs(optimal_lambda) < 0.1) {
  abalone <- abalone %>% mutate(Age_bc = log(Age))
  bc_transform <- "log"
} else {
  abalone <- abalone %>% mutate(Age_bc = (Age^optimal_lambda - 1) / optimal_lambda)
  bc_transform <- paste0("power_", round(optimal_lambda, 2))
}
cat("Box-Cox transformation applied:", bc_transform, "\n")
```

# Part 5: Exploratory Data Analysis

## Age Distribution

```{r eda-age}
eda1 <- abalone %>%
  ggplot(aes(x = Age)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean(Age)), color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(aes(xintercept = median(Age)), color = "orange", linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of Abalone Age",
       subtitle = paste("Mean:", round(mean(abalone$Age), 1), "| Median:", median(abalone$Age)),
       x = "Age (years)", y = "Count") +
  theme_minimal()

eda2 <- abalone %>%
  ggplot(aes(x = Sex, y = Age, fill = Sex)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("F" = "pink", "I" = "lightgreen", "M" = "lightblue")) +
  labs(title = "Age Distribution by Sex", x = "Sex", y = "Age (years)") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(eda1, eda2, ncol = 2)
```

## Weight vs Age

```{r eda-weight}
eda3 <- abalone %>%
  ggplot(aes(x = Weight, y = Age, color = Sex)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(values = c("F" = "red", "I" = "green", "M" = "blue")) +
  labs(title = "Age vs Total Weight", x = "Weight (grams)", y = "Age (years)") +
  theme_minimal()

eda4 <- abalone %>%
  ggplot(aes(x = `Shell Weight`, y = Age, color = Sex)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(values = c("F" = "red", "I" = "green", "M" = "blue")) +
  labs(title = "Age vs Shell Weight", x = "Shell Weight (grams)", y = "Age (years)") +
  theme_minimal()

grid.arrange(eda3, eda4, ncol = 2)
```

## Size and Dimensions

```{r eda-size}
eda5 <- abalone %>%
  ggplot(aes(x = size_mean, y = Age, color = Sex)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(values = c("F" = "red", "I" = "green", "M" = "blue")) +
  labs(title = "Age vs Size Mean", x = "Size Mean", y = "Age (years)") +
  theme_minimal()

eda6 <- abalone %>%
  dplyr::select(Length, Diameter, Height) %>%
  pivot_longer(everything(), names_to = "Dimension", values_to = "Value") %>%
  ggplot(aes(x = Value, fill = Dimension)) +
  geom_histogram(bins = 30, alpha = 0.7, color = "white") +
  facet_wrap(~Dimension, scales = "free") +
  labs(title = "Distribution of Physical Dimensions", x = "Value (inches)", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(eda5, eda6, ncol = 2)
```

## Log-Log Relationship

```{r eda-loglog}
abalone %>%
  ggplot(aes(x = log_weight, y = log(Age), color = Sex)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(values = c("F" = "red", "I" = "green", "M" = "blue")) +
  labs(title = "log(Age) vs log(Weight) - Linearized Relationship",
       x = "log(Weight)", y = "log(Age)") +
  theme_minimal()
```

## Correlation Matrix

```{r correlation}
cor_data <- abalone %>%
  dplyr::select(Age, Length, Diameter, Height, Weight, 
                `Shucked Weight`, `Viscera Weight`, `Shell Weight`,
                size_mean, shell_thickness_proxy, density) %>%
  cor()

corrplot(cor_data, method = "color", type = "upper", 
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix", mar = c(0, 0, 2, 0))
```

# Part 6: Cross-Validation Setup

```{r cv-setup}
set.seed(123)

train_control <- trainControl(
  method = "cv",
  number = 10,
  savePredictions = "final"
)

calc_metrics <- function(model_cv, transform = "none") {
  preds_raw <- model_cv$pred$pred
  actual_raw <- model_cv$pred$obs
  
  if (transform == "log") {
    preds <- exp(preds_raw)
    actual <- exp(actual_raw)
  } else if (transform == "boxcox") {
    if (abs(optimal_lambda) < 0.1) {
      preds <- exp(preds_raw)
      actual <- exp(actual_raw)
    } else {
      preds <- (preds_raw * optimal_lambda + 1)^(1/optimal_lambda)
      actual <- (actual_raw * optimal_lambda + 1)^(1/optimal_lambda)
    }
  } else {
    preds <- preds_raw
    actual <- actual_raw
  }
  
  list(
    MAE = mean(abs(actual - preds)),
    MAPE = mean(abs((actual - preds) / actual)) * 100,
    RMSE = sqrt(mean((actual - preds)^2)),
    R2 = 1 - sum((actual - preds)^2) / sum((actual - mean(actual))^2)
  )
}
```

# Part 7: Standard Linear Models (M1-M11)

```{r models-standard}
m1 <- train(Age ~ Sex + Length + Diameter + Height + Weight + 
              `Shucked Weight` + `Viscera Weight` + `Shell Weight`,
            data = abalone, method = "lm", trControl = train_control)

m2 <- train(Age ~ is_infant + size_mean + Weight + shell_thickness_proxy,
            data = abalone, method = "lm", trControl = train_control)

m3 <- train(Age ~ is_infant + size_mean + Weight + shell_thickness_proxy + 
              shell_ratio + shucked_ratio,
            data = abalone, method = "lm", trControl = train_control)

m4 <- train(Age ~ is_infant + size_mean + Weight + shell_thickness_proxy + 
              shell_ratio + shucked_ratio + is_slow_growth,
            data = abalone, method = "lm", trControl = train_control)

m5 <- train(Age ~ is_infant + size_mean + I(size_mean^2) + Weight + 
              shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth,
            data = abalone, method = "lm", trControl = train_control)

m6 <- train(Age ~ is_infant + size_mean + I(size_mean^2) + Weight + 
              shell_thickness_proxy + shell_ratio + shucked_ratio + 
              is_slow_growth + size_mean:Weight,
            data = abalone, method = "lm", trControl = train_control)

m7 <- train(log(Age) ~ is_infant + size_mean + Weight + shell_thickness_proxy + 
              shell_ratio + shucked_ratio + is_slow_growth,
            data = abalone, method = "lm", trControl = train_control)

m8 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + Weight + 
              shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth,
            data = abalone, method = "lm", trControl = train_control)

m9 <- train(log(Age) ~ is_infant + log(size_mean) + Weight + 
              shell_thickness_proxy + shell_ratio + is_slow_growth,
            data = abalone, method = "lm", trControl = train_control)

m10 <- train(Age ~ is_infant + shell_thickness_proxy + Height + size_mean + 
               `Shell Weight` + density,
             data = abalone, method = "lm", trControl = train_control)

m11 <- train(Age ~ is_infant + size_mean + I(size_mean^2) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + 
               viscera_ratio + density + volume_ellipsoid + is_slow_growth +
               height_diameter_ratio,
             data = abalone, method = "lm", trControl = train_control)
```

# Part 8: Interaction Models (M12-M15)

```{r models-interaction}
m12 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
               size_shell_interaction + weight_infant_interaction,
             data = abalone, method = "lm", trControl = train_control)

m13 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + I(size_mean^3) + 
               Weight + shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth,
             data = abalone, method = "lm", trControl = train_control)

m14 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
               density_size_interaction,
             data = abalone, method = "lm", trControl = train_control)

m15 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
               size_shell_interaction + size_shucked_interaction + 
               weight_infant_interaction + shell_infant_interaction,
             data = abalone, method = "lm", trControl = train_control)
```

# Part 9: Regularized Models (M16-M18)

```{r models-regularized}
formula_reg <- log(Age) ~ is_infant + size_mean + size_mean_squared + Weight + 
  shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
  size_shell_interaction + size_shucked_interaction + weight_infant_interaction +
  density + volume_ellipsoid + height_diameter_ratio

ridge_grid <- expand.grid(alpha = 0, lambda = 10^seq(-4, 0, length = 20))
m16 <- train(formula_reg, data = abalone, method = "glmnet",
             trControl = train_control, tuneGrid = ridge_grid)

lasso_grid <- expand.grid(alpha = 1, lambda = 10^seq(-4, 0, length = 20))
m17 <- train(formula_reg, data = abalone, method = "glmnet",
             trControl = train_control, tuneGrid = lasso_grid)

enet_grid <- expand.grid(alpha = 0.5, lambda = 10^seq(-4, 0, length = 20))
m18 <- train(formula_reg, data = abalone, method = "glmnet",
             trControl = train_control, tuneGrid = enet_grid)
```

# Part 10: Nonlinearity Fix Models (M19-M27)

```{r models-nonlinearity}
m19 <- train(log(Age) ~ is_infant + size_mean + log_weight + log_shell_weight + 
               shell_ratio + shucked_ratio + is_slow_growth,
             data = abalone, method = "lm", trControl = train_control)

m20 <- train(log(Age) ~ is_infant + log_size_mean + log_weight + log_shell_weight + 
               log_shucked_weight + is_slow_growth,
             data = abalone, method = "lm", trControl = train_control)

m21 <- train(log(Age) ~ is_infant + ns(size_mean, df = 3) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth,
             data = abalone, method = "lm", trControl = train_control)

m22 <- train(log(Age) ~ is_infant + ns(size_mean, df = 4) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth,
             data = abalone, method = "lm", trControl = train_control)

m23 <- train(log(Age) ~ is_infant + ns(size_mean, df = 3) + ns(Weight, df = 3) + 
               shell_ratio + shucked_ratio + is_slow_growth,
             data = abalone, method = "lm", trControl = train_control)

m24 <- train(Age_bc ~ is_infant + size_mean + I(size_mean^2) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth,
             data = abalone, method = "lm", trControl = train_control)

m25 <- train(log(Age) ~ is_infant + ns(size_mean, df = 3) + log_weight + 
               log_shell_weight + shell_ratio + shucked_ratio + is_slow_growth +
               weight_infant_interaction,
             data = abalone, method = "lm", trControl = train_control)

m26 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + I(size_mean^3) + 
               I(size_mean^4) + Weight + shell_thickness_proxy + shell_ratio + 
               shucked_ratio + is_slow_growth,
             data = abalone, method = "lm", trControl = train_control)

m27 <- train(log(Age) ~ is_infant + sqrt_size_mean + sqrt_weight + sqrt_shell_weight + 
               shell_ratio + shucked_ratio + is_slow_growth,
             data = abalone, method = "lm", trControl = train_control)
```

# Part 11: Enhanced Interaction Models (M28-M38)

```{r models-enhanced}
m28 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
               size_shell_interaction + size_shucked_interaction + 
               weight_infant_interaction + shell_infant_interaction +
               density_infant_interaction,
             data = abalone, method = "lm", trControl = train_control)

m29 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
               size_shell_interaction + size_shucked_interaction + 
               weight_infant_interaction + shell_infant_interaction +
               weight_shell_interaction + weight_shucked_interaction,
             data = abalone, method = "lm", trControl = train_control)

m30 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
               size_shell_interaction + size_shucked_interaction + 
               weight_infant_interaction + shell_infant_interaction +
               slow_growth_size,
             data = abalone, method = "lm", trControl = train_control)

m31 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
               size_shell_interaction + size_shucked_interaction + 
               weight_infant_interaction + shell_infant_interaction +
               infant_size_squared,
             data = abalone, method = "lm", trControl = train_control)

m32 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + I(size_mean^3) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
               size_shell_interaction + size_shucked_interaction + 
               weight_infant_interaction + shell_infant_interaction,
             data = abalone, method = "lm", trControl = train_control)

m33 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
               size_shell_interaction + size_shucked_interaction + 
               weight_infant_interaction + shell_infant_interaction +
               density_infant_interaction + slow_growth_size + infant_size_squared,
             data = abalone, method = "lm", trControl = train_control)

m34 <- train(log(Age) ~ is_infant + ns(size_mean, df = 3) + Weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
               size_shell_interaction + size_shucked_interaction + 
               weight_infant_interaction + shell_infant_interaction,
             data = abalone, method = "lm", trControl = train_control)

m35 <- train(log(Age) ~ is_infant + size_mean + I(size_mean^2) + log_weight + 
               shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
               size_shell_interaction + size_shucked_interaction + 
               weight_infant_interaction + shell_infant_interaction,
             data = abalone, method = "lm", trControl = train_control)

# Enhanced regularized models
formula_enhanced <- log(Age) ~ is_infant + size_mean + size_mean_squared + size_mean_cubed +
  Weight + shell_thickness_proxy + shell_ratio + shucked_ratio + viscera_ratio +
  is_slow_growth + density + height_diameter_ratio +
  size_shell_interaction + size_shucked_interaction + size_viscera_interaction +
  weight_infant_interaction + shell_infant_interaction + density_infant_interaction +
  slow_growth_size + infant_size_squared + weight_shell_interaction

ridge_grid_enh <- expand.grid(alpha = 0, lambda = 10^seq(-4, 0, length = 20))
m36 <- train(formula_enhanced, data = abalone, method = "glmnet",
             trControl = train_control, tuneGrid = ridge_grid_enh)

lasso_grid_enh <- expand.grid(alpha = 1, lambda = 10^seq(-4, 0, length = 20))
m37 <- train(formula_enhanced, data = abalone, method = "glmnet",
             trControl = train_control, tuneGrid = lasso_grid_enh)

enet_grid_enh <- expand.grid(alpha = 0.5, lambda = 10^seq(-4, 0, length = 20))
m38 <- train(formula_enhanced, data = abalone, method = "glmnet",
             trControl = train_control, tuneGrid = enet_grid_enh)
```

# Part 12: Compile Results

```{r compile-results}
metrics_list <- list(
  calc_metrics(m1, "none"), calc_metrics(m2, "none"), calc_metrics(m3, "none"),
  calc_metrics(m4, "none"), calc_metrics(m5, "none"), calc_metrics(m6, "none"),
  calc_metrics(m7, "log"), calc_metrics(m8, "log"), calc_metrics(m9, "log"),
  calc_metrics(m10, "none"), calc_metrics(m11, "none"),
  calc_metrics(m12, "log"), calc_metrics(m13, "log"), calc_metrics(m14, "log"),
  calc_metrics(m15, "log"), calc_metrics(m16, "log"), calc_metrics(m17, "log"),
  calc_metrics(m18, "log"),
  calc_metrics(m19, "log"), calc_metrics(m20, "log"), calc_metrics(m21, "log"),
  calc_metrics(m22, "log"), calc_metrics(m23, "log"), calc_metrics(m24, "boxcox"),
  calc_metrics(m25, "log"), calc_metrics(m26, "log"), calc_metrics(m27, "log"),
  calc_metrics(m28, "log"), calc_metrics(m29, "log"), calc_metrics(m30, "log"),
  calc_metrics(m31, "log"), calc_metrics(m32, "log"), calc_metrics(m33, "log"),
  calc_metrics(m34, "log"), calc_metrics(m35, "log"), calc_metrics(m36, "log"),
  calc_metrics(m37, "log"), calc_metrics(m38, "log")
)

model_comparison <- tibble(
  Model = c("M1_Baseline", "M2_Simple", "M3_Ratios", "M4_BioFlag", 
            "M5_Polynomial", "M6_Interaction", "M7_Log", "M8_Log_Poly",
            "M9_Log_Log", "M10_TopCorr", "M11_KitchenSink",
            "M12_Log_MoreInt", "M13_Log_Cubic", "M14_Log_Density", "M15_Log_FullInt",
            "M16_Ridge", "M17_Lasso", "M18_ElasticNet",
            "M19_LogWeights", "M20_AllLog", "M21_Spline3", "M22_Spline4",
            "M23_Spline_Both", "M24_BoxCox", "M25_Spline_LogWt", 
            "M26_Quartic", "M27_Sqrt",
            "M28_DensityInt", "M29_WeightRatioInt", "M30_SlowGrowthInt",
            "M31_InfantSizeSq", "M32_CubicInt", "M33_KitchenSinkInt",
            "M34_SplineInt", "M35_LogWeightInt", "M36_Ridge_Enh",
            "M37_Lasso_Enh", "M38_ElasticNet_Enh"),
  MAE = round(map_dbl(metrics_list, "MAE"), 4),
  MAPE = round(map_dbl(metrics_list, "MAPE"), 2),
  RMSE = round(map_dbl(metrics_list, "RMSE"), 4),
  R2 = round(map_dbl(metrics_list, "R2"), 4),
  Type = c(rep("Standard LM", 11), rep("LM + Interactions", 4), 
           rep("Regularized LM", 3), rep("Nonlinearity Fix", 9),
           rep("Enhanced Interactions", 8), rep("Enhanced Regularized", 3))
) %>%
  arrange(MAE)
```

# Part 13: Ensemble Models

```{r ensembles}
get_cv_preds <- function(model_cv, transform = "none") {
  preds <- model_cv$pred %>% arrange(rowIndex)
  if (transform == "log") {
    return(exp(preds$pred))
  } else if (transform == "boxcox") {
    if (abs(optimal_lambda) < 0.1) {
      return(exp(preds$pred))
    } else {
      return((preds$pred * optimal_lambda + 1)^(1/optimal_lambda))
    }
  } else {
    return(preds$pred)
  }
}

all_preds <- tibble(
  actual = abalone$Age,
  M7_Log = get_cv_preds(m7, "log"),
  M8_Log_Poly = get_cv_preds(m8, "log"),
  M15_Log_FullInt = get_cv_preds(m15, "log"),
  M17_Lasso = get_cv_preds(m17, "log"),
  M18_ElasticNet = get_cv_preds(m18, "log"),
  M21_Spline3 = get_cv_preds(m21, "log"),
  M32_CubicInt = get_cv_preds(m32, "log"),
  M33_KitchenSinkInt = get_cv_preds(m33, "log"),
  M37_Lasso_Enh = get_cv_preds(m37, "log"),
  M38_ElasticNet_Enh = get_cv_preds(m38, "log")
)

# E1: Top 3 original models
ensemble_top3_orig <- rowMeans(all_preds[, c("M8_Log_Poly", "M7_Log", "M15_Log_FullInt")])

# E2: Best nonlinearity + interactions
ensemble_nonlinear <- rowMeans(all_preds[, c("M21_Spline3", "M32_CubicInt", "M33_KitchenSinkInt")])

# E3: Mix of best from each type
ensemble_mix <- rowMeans(all_preds[, c("M15_Log_FullInt", "M37_Lasso_Enh", "M38_ElasticNet_Enh")])

# E4: All top models
ensemble_all <- rowMeans(all_preds[, -1])

# E5: Weighted ensemble
top3_mae <- model_comparison %>% slice(1:3) %>% pull(MAE)
weights <- (1/top3_mae) / sum(1/top3_mae)
ensemble_weighted <- weights[1] * all_preds$M15_Log_FullInt + 
                     weights[2] * all_preds$M37_Lasso_Enh + 
                     weights[3] * all_preds$M38_ElasticNet_Enh

calc_ensemble_metrics <- function(preds, actual) {
  list(
    MAE = mean(abs(actual - preds)),
    MAPE = mean(abs((actual - preds) / actual)) * 100,
    RMSE = sqrt(mean((actual - preds)^2)),
    R2 = 1 - sum((actual - preds)^2) / sum((actual - mean(actual))^2)
  )
}

e1_metrics <- calc_ensemble_metrics(ensemble_top3_orig, all_preds$actual)
e2_metrics <- calc_ensemble_metrics(ensemble_nonlinear, all_preds$actual)
e3_metrics <- calc_ensemble_metrics(ensemble_mix, all_preds$actual)
e4_metrics <- calc_ensemble_metrics(ensemble_all, all_preds$actual)
e5_metrics <- calc_ensemble_metrics(ensemble_weighted, all_preds$actual)

ensemble_results <- tibble(
  Model = c("E1_Top3_Orig", "E2_Nonlinear", "E3_Mix", "E4_All", "E5_Weighted"),
  MAE = round(c(e1_metrics$MAE, e2_metrics$MAE, e3_metrics$MAE, e4_metrics$MAE, e5_metrics$MAE), 4),
  MAPE = round(c(e1_metrics$MAPE, e2_metrics$MAPE, e3_metrics$MAPE, e4_metrics$MAPE, e5_metrics$MAPE), 2),
  RMSE = round(c(e1_metrics$RMSE, e2_metrics$RMSE, e3_metrics$RMSE, e4_metrics$RMSE, e5_metrics$RMSE), 4),
  R2 = round(c(e1_metrics$R2, e2_metrics$R2, e3_metrics$R2, e4_metrics$R2, e5_metrics$R2), 4),
  Type = "Ensemble"
)

final_comparison <- bind_rows(model_comparison, ensemble_results) %>%
  arrange(MAE)
```

# Part 14: Results

## All Models Comparison

```{r results-table}
knitr::kable(final_comparison, caption = "All 43 Models Ranked by MAE")
```

## Top 10 Models

```{r top10}
final_comparison %>% 
  slice(1:10) %>%
  knitr::kable(caption = "Top 10 Models")
```

## Best by Type

```{r best-by-type}
final_comparison %>% 
  group_by(Type) %>% 
  slice_min(MAE, n = 1) %>%
  knitr::kable(caption = "Best Model by Type")
```

# Part 15: Prediction Plots

```{r prediction-plots, fig.height=12}
best_preds <- all_preds %>%
  dplyr::select(actual, M15_Log_FullInt) %>%
  rename(predicted = M15_Log_FullInt) %>%
  mutate(residual = actual - predicted)

pred_plot1 <- best_preds %>%
  ggplot(aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen", alpha = 0.3) +
  labs(title = "Actual vs Predicted Age",
       subtitle = paste("MAE =", round(mean(abs(best_preds$residual)), 3)),
       x = "Actual Age (years)", y = "Predicted Age (years)") +
  theme_minimal() +
  coord_equal()

pred_plot2 <- best_preds %>%
  ggplot(aes(x = predicted, y = residual)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  geom_smooth(method = "loess", se = TRUE, color = "darkgreen", alpha = 0.3) +
  labs(title = "Residuals vs Predicted Age",
       x = "Predicted Age (years)", y = "Residual") +
  theme_minimal()

pred_plot3 <- best_preds %>%
  ggplot(aes(x = residual)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of Residuals",
       subtitle = paste("Mean:", round(mean(best_preds$residual), 3),
                        "| SD:", round(sd(best_preds$residual), 3)),
       x = "Residual (years)", y = "Count") +
  theme_minimal()

pred_plot4 <- abalone %>%
  mutate(predicted = best_preds$predicted) %>%
  ggplot(aes(x = Age, y = predicted, color = Sex)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  scale_color_manual(values = c("F" = "red", "I" = "green", "M" = "blue")) +
  facet_wrap(~Sex) +
  labs(title = "Actual vs Predicted Age by Sex",
       x = "Actual Age (years)", y = "Predicted Age (years)") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(pred_plot1, pred_plot2, pred_plot3, pred_plot4, ncol = 2)
```

# Part 16: Model Comparison Plots

```{r model-comparison-plots, fig.height=16}
p1 <- final_comparison %>%
  mutate(Model = fct_reorder(Model, MAE, .desc = TRUE)) %>%
  ggplot(aes(x = MAE, y = Model, fill = Type)) +
  geom_col() +
  geom_vline(xintercept = min(final_comparison$MAE), 
             color = "red", linetype = "dashed", alpha = 0.7) +
  scale_fill_manual(values = c("Standard LM" = "steelblue", 
                               "LM + Interactions" = "darkgreen",
                               "Regularized LM" = "purple",
                               "Nonlinearity Fix" = "coral",
                               "Enhanced Interactions" = "gold",
                               "Enhanced Regularized" = "darkred",
                               "Ensemble" = "orange")) +
  labs(title = "All Models: MAE (Lower is Better)", x = "MAE (years)", y = NULL) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))

p2 <- final_comparison %>%
  mutate(Model = fct_reorder(Model, RMSE, .desc = TRUE)) %>%
  ggplot(aes(x = RMSE, y = Model, fill = Type)) +
  geom_col() +
  scale_fill_manual(values = c("Standard LM" = "steelblue", 
                               "LM + Interactions" = "darkgreen",
                               "Regularized LM" = "purple",
                               "Nonlinearity Fix" = "coral",
                               "Enhanced Interactions" = "gold",
                               "Enhanced Regularized" = "darkred",
                               "Ensemble" = "orange")) +
  labs(title = "All Models: RMSE (Lower is Better)", x = "RMSE (years)", y = NULL) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))

p3 <- final_comparison %>%
  mutate(Model = fct_reorder(Model, R2)) %>%
  ggplot(aes(x = R2, y = Model, fill = Type)) +
  geom_col() +
  scale_fill_manual(values = c("Standard LM" = "steelblue", 
                               "LM + Interactions" = "darkgreen",
                               "Regularized LM" = "purple",
                               "Nonlinearity Fix" = "coral",
                               "Enhanced Interactions" = "gold",
                               "Enhanced Regularized" = "darkred",
                               "Ensemble" = "orange")) +
  labs(title = "All Models: R² (Higher is Better)", x = "R²", y = NULL) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))

p4 <- final_comparison %>%
  ggplot(aes(x = MAE, y = R2, color = Type)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("Standard LM" = "steelblue", 
                                "LM + Interactions" = "darkgreen",
                                "Regularized LM" = "purple",
                                "Nonlinearity Fix" = "coral",
                                "Enhanced Interactions" = "gold",
                                "Enhanced Regularized" = "darkred",
                                "Ensemble" = "orange")) +
  labs(title = "MAE vs R²", x = "MAE (years)", y = "R²") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

# Part 17: Competition Predictions

```{r competition-predictions}
best_model_name <- final_comparison %>% slice(1) %>% pull(Model)
best_mae <- final_comparison %>% slice(1) %>% pull(MAE)
cat("Best Model:", best_model_name, "with MAE =", best_mae, "\n")

# Train final models on ALL data
final_m15 <- lm(log(Age) ~ is_infant + size_mean + I(size_mean^2) + Weight + 
                  shell_thickness_proxy + shell_ratio + shucked_ratio + is_slow_growth +
                  size_shell_interaction + size_shucked_interaction + 
                  weight_infant_interaction + shell_infant_interaction,
                data = abalone)

final_m37 <- lm(log(Age) ~ is_infant + size_mean + I(size_mean^2) + I(size_mean^3) + Weight + 
                  shell_thickness_proxy + shell_ratio + shucked_ratio + viscera_ratio +
                  is_slow_growth + density + height_diameter_ratio +
                  size_shell_interaction + size_shucked_interaction + size_viscera_interaction +
                  weight_infant_interaction + shell_infant_interaction + density_infant_interaction +
                  slow_growth_size + infant_size_squared + weight_shell_interaction,
                data = abalone)

final_m38 <- lm(log(Age) ~ is_infant + size_mean + I(size_mean^2) + I(size_mean^3) + Weight + 
                  shell_thickness_proxy + shell_ratio + shucked_ratio + viscera_ratio +
                  is_slow_growth + density + height_diameter_ratio +
                  size_shell_interaction + size_shucked_interaction + size_viscera_interaction +
                  weight_infant_interaction + shell_infant_interaction + density_infant_interaction +
                  slow_growth_size + infant_size_squared + weight_shell_interaction,
                data = abalone)
```

```{r load-competition}
competition <- read_csv("~/Library/CloudStorage/Box-Box/Data Science/1. MSDS - Doing Data Science (Bivin Sadler)/Unit 14 and 15 (Group Project)/competition.csv", show_col_types = FALSE)

# Apply same preprocessing
competition <- competition %>%
  mutate(
    component_sum = `Shucked Weight` + `Viscera Weight` + `Shell Weight`,
    Weight = if_else(component_sum > Weight, component_sum, Weight)
  ) %>%
  dplyr::select(-component_sum) %>%
  left_join(medians_by_sex, by = "Sex") %>%
  mutate(
    Length = if_else(Length == 0 | is.na(Length), median_Length, Length),
    Diameter = if_else(Diameter == 0 | is.na(Diameter), median_Diameter, Diameter),
    Height = if_else(Height == 0 | is.na(Height), median_Height, Height)
  ) %>%
  dplyr::select(-median_Length, -median_Diameter, -median_Height) %>%
  mutate(
    size_mean = (Length * Diameter * Height)^(1/3),
    size_mean_squared = size_mean^2,
    size_mean_cubed = size_mean^3,
    volume_ellipsoid = (4/3) * pi * (Length/2) * (Diameter/2) * (Height/2),
    density = Weight / volume_ellipsoid,
    surface_area = 4 * pi * ((Length * Diameter)^1.6075 / 3)^(1/1.6075),
    shell_thickness_proxy = `Shell Weight` / surface_area,
    shucked_ratio = `Shucked Weight` / Weight,
    viscera_ratio = `Viscera Weight` / Weight,
    shell_ratio = `Shell Weight` / Weight,
    height_diameter_ratio = Height / Diameter,
    is_infant = if_else(Sex == "I", 1L, 0L),
    is_slow_growth = 0L,
    size_shell_interaction = size_mean * shell_ratio,
    size_shucked_interaction = size_mean * shucked_ratio,
    size_viscera_interaction = size_mean * viscera_ratio,
    weight_infant_interaction = Weight * is_infant,
    shell_infant_interaction = shell_ratio * is_infant,
    density_infant_interaction = density * is_infant,
    slow_growth_size = is_slow_growth * size_mean,
    infant_size_squared = is_infant * size_mean_squared,
    weight_shell_interaction = Weight * shell_ratio,
    Sex = factor(Sex)
  ) %>%
  mutate(across(where(is.numeric), ~if_else(is.infinite(.), NA_real_, .))) %>%
  mutate(across(where(is.numeric), ~if_else(is.na(.), median(., na.rm = TRUE), .)))

# Generate ensemble predictions
pred_m15 <- exp(predict(final_m15, newdata = competition))
pred_m37 <- exp(predict(final_m37, newdata = competition))
pred_m38 <- exp(predict(final_m38, newdata = competition))

predictions <- competition %>%
  mutate(age = (pred_m15 + pred_m37 + pred_m38) / 3) %>%
  dplyr::select(id, age)

# Save predictions
write_csv(predictions, "abalone_predictions.csv")
```

## Prediction Summary

```{r prediction-summary}
cat("Competition Predictions Saved\n\n")
head(predictions, 10)
summary(predictions$age)
```

# Part 18: Final Summary

```{r final-summary}
cat("============================================================\n")
cat("                    FINAL SUMMARY                           \n")
cat("============================================================\n\n")
cat("Total Models Tested: 43 (38 single + 5 ensembles)\n\n")
cat("Model Categories:\n")
cat("  - Standard LM:           11 models (M1-M11)\n")
cat("  - LM + Interactions:      4 models (M12-M15)\n")
cat("  - Regularized LM:         3 models (M16-M18)\n")
cat("  - Nonlinearity Fix:       9 models (M19-M27)\n")
cat("  - Enhanced Interactions:  8 models (M28-M35)\n")
cat("  - Enhanced Regularized:   3 models (M36-M38)\n")
cat("  - Ensembles:              5 models (E1-E5)\n\n")
cat("Best Model:", best_model_name, "\n")
cat("Best MAE:", best_mae, "years\n")
cat("============================================================\n")
```
